---
title: "Assignment 1"
author: 'Name: Jash Prakash Rana, ID: 22222806'
date: "16th Feb, 2022"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Part 2
## Implementation of Part 1 using ggplot2 in RStudio, along with some modifications explained below.
</br>

### 1. <u>Importing the data</u>
```{r}
cpi_data <- read.csv("Counsumer_Price_Index.csv")

#Ref: https://www.geeksforgeeks.org/delete-rows-with-empty-cells-from-excel-using-r/
cpi_data <- cpi_data[!apply(is.na(cpi_data) | cpi_data == "", 1, all),]
```
### 2. <u>Creating a plotting data using `dplyr::filter` package.</u>
* Filtering the data according to Dec 2022
* Changing the column name of Percentage to `Per_2022_12`
* Removing the Time and Frequency columns
* Changing 'Total_Minus_Food_Energy' to `TMSE`
* Factoring `Subject` column to show Total, Food and TMFE in that level.
```{r message=FALSE, warning=FALSE}
library(dplyr)

sub_req = c("Total","Food","Total_Minus_Food_Energy")
loc_req = c("IRL","GBR","USA","AUT","MEX","FRA","ESP","CAN","G-7")

plot_data <- cpi_data%>%
  dplyr::filter(Time == "2022-12", 
         Subject %in% sub_req,
         Location %in% loc_req)
colnames(plot_data)[5] <- "Per_2022_12"
plot_data <- subset(plot_data, select = -c(Time, Frequency))
plot_data$Subject <- gsub("Total_Minus_Food_Energy","TMFE",plot_data$Subject)
plot_data$Subject <- factor(plot_data$Subject, levels = c("Total", "Food", "TMFE"))
head(plot_data)
```
### 3. <u>Creating a Dot Plot</u></br>
The dot plot was created in focusing the Location `IRL` into the foreground, while the other countries and G-7 into the background. The major change that we see from part 1 is the focus on out country `IRL`, and instead of the Location in the X-axis, I decided to use Subject in the X-axis as that is something which can be grouped/reordered by the Percentage (Per_2022_12) attribute. While I also have coloured the Group of Nations' shape to a blue shade and our country by the pink shade, I decided to keep all the other countries by the grey colour as shown in the Legend, and put a caption which is displayed directly below the Legend which talks about the countries included in the 'Other' section. I have given shape 21 (circle with colour fill) to IRL while we have given shape 4 (X or cross) to all the other countries. I have disabled only the minor gridline at X-axis as our data is categorical and not a continuous value. I have also changed the Y-axis title to rotate it, so that we can see it without any neck tilt. And finally, the legend was taken below the X-axis, in the center, to save space and also to make the graph bigger in size by the width.</br>

I agree that the dot plot is cluttered from Part 1 but I found this approach to be a better way to show as we can finally see the pattern showing where our country lies in the Location data, while ranking it in each different measure of Subject column. We can make an easy comparison this way between our country and the others this way according to the measures, and it becomes clear on how much change did our country have compared to other. As said in Part 1, the better approach in this scenario was to use bar charts and as we progress, we will see why that approach clears many things visually and how it is easy to understand our data using that approach.
```{r}
library(ggplot2)

shapes <- c(4, 21)
Others <- c("GBR","USA","AUT","MEX","FRA","ESP","CAN")

dotplot <- ggplot(plot_data, aes(x = Subject, y = Per_2022_12, 
                                 group = -Per_2022_12, 
                                 colour = Location,
                                 fill = Location,
                                 alpha = Location == "IRL",
                                 shape = Location == "IRL")) +
  geom_point(size = 3.2) +
  scale_shape_manual(values = c(4, 21)) +
  scale_fill_manual(values = c("IRL" = "#D81B60", "G-7" = "#1E88E5", Others = "lightgrey"))+ 
  scale_colour_manual("Countries/Group of Nation", 
                    values = c("IRL" = "#D81B60", "G-7" = "#1E88E5", Others = "lightgrey"))+
  scale_alpha_manual(values = c(0.7, 1), guide = "none") +
  labs(title = "Consumer Price Index (CPI) Comparison", 
       subtitle = "Change in December 2022 since December 2021",
       caption = "Other Countries = USA, AUT, MEX, FRA, ESP & CAN") +
  xlab("Countries/Group of Nations") +
  ylab("Change in \nCPI (%)") + 
  guides(
    colour = guide_legend("Countries/Group of Nation"),
    fill = guide_legend("Countries/Group of Nation"),
    shape = "none") +
  scale_y_continuous(breaks = seq(0,20, by = 4), limits = c(2,18)) +
  theme(panel.background = element_rect(fill = "white"),
        panel.grid.major.x = element_line(size = 0.1, linetype = "dashed",
                                        colour = 'lightgrey'),
        panel.grid.major.y = element_line(size = 0.12, linetype = "solid",
                                          colour = 'lightgrey'),
        panel.grid.minor.y = element_line(size = 0.068, linetype = "dotted",
                                          colour = 'lightgrey'),
        axis.ticks = element_blank(),
        axis.text = element_text(size = 9),
        axis.title = element_text(size = 11),
        axis.title.y = element_text(size = 11,
                                 angle = 0,
                                 vjust=0.5,
                                 hjust=0,
                                 margin = margin(r=4)),

        legend.title = element_text(face="italic", size=11),
        legend.text = element_text(face = "italic", size = 9),
        legend.box.background = element_rect(fill = 'lightgrey', colour = 'lightgrey'),
        legend.key = element_rect(fill = "white"),
        legend.position = "bottom",
        legend.direction = "horizontal",
        legend.justification = "center",
        legend.box.margin = margin(0.5, 0.5, 0.5, 0.5),

        plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(face = 'italic', size = 12))

dotplot
  
```
\vspace{5mm}

### 4. <u>Creating a Bar Chart</u></br>
Using the same colour scale we used in the dot plot above, our country `IRL` was given a pink shade, G-7 the blue shade, and the rest with grey shade. The alpha values were modified to keep our country at the forefront and solid in the chart, while keep everything else in the background as light as possible. The bar chart is grouping descending by the Percentage (Per_2022_12) values. Major change from Part 1 is that I am using Subject column in the X-axis as opposed to Location column. I have kept only the Major and Minor Y-axis gridlines to keep this chart minimal. To keep our country in foreground, the alpha value was set lower for other countries. I decided to use Subject in the X-axis as that is something which can be grouped/reordered by the Percentage (Per_2022_12) attribute. Also the captions mentions the other countries marked by the grey colour in the chart. I have also changed the Y-axis title to rotate it, so that we can see it without any neck tilt. And finally, the legend was taken below the X-axis, in the center, to save space and also to make the graph bigger in size by the width.</br>

This is a much different graph from Part 1 as it focuses on our country when all the measures were separated on X-axis to show descending values of all the countries when compared to our country. This is definitely a better chart in terms of pattern-finding and appeal, while this becomes very easy to read chart even for a non-related person of the field who can infer basic things from this graph, and that was the goal to plotting a graph in Part 2. This approach is way beneficial if compared even to the dot plot as this shows better readability and less clutter.
```{r}
library(RColorBrewer)
library(ggplot2)
library(dplyr)

Others <- c("GBR","USA","AUT","MEX","FRA","ESP","CAN")

barplot_new <- ggplot(plot_data, aes(x = Subject, y = Per_2022_12)) +
  geom_col(position = "dodge2", width = 0.9,
           aes(group = -Per_2022_12, fill = Location, alpha = Location == "IRL"))+
  scale_fill_manual("Countries/Group of Nation", 
                    values = c("IRL" = "#D81B60", "G-7" = "#1E88E5", Others = "grey"), aes(group = Per_2022_12))+
  scale_alpha_manual(values = c(0.38, 1), guide = "none")+
  labs(title = "Consumer Price Index (CPI) Comparison", 
       subtitle = "Change in December 2022 since December 2021",
       caption = "Other Countries = USA, AUT, MEX, FRA, ESP & CAN") +
  xlab("Subject") +
  ylab("Change in \nCPI (%)") +
  scale_y_continuous(breaks = seq(0,20, by = 4), limits = c(0,18), expand = c(0,0)) +
  theme(panel.background = element_rect(fill = "white"),
        panel.grid.major.y = element_line(size = 0.12, linetype = "solid",
                                          colour = 'lightgrey'),
        panel.grid.minor.y = element_line(size = 0.068, linetype = "dashed",
                                          colour = 'lightgrey'),
        axis.ticks = element_blank(),
        axis.text = element_text(size = 9),
        axis.title = element_text(size = 11),
        axis.title.y = element_text(size = 11, 
                                 angle = 0,
                                 vjust=0.5,
                                 hjust=0,
                                 margin = margin(r=4)),
        
        legend.title = element_text(face="italic", size=11),
        legend.text = element_text(face="italic", size = 9),
        legend.box.background = element_rect(fill = 'lightgrey', colour = 'lightgrey'),
        legend.key = element_rect(fill = "white"),
        legend.position = "bottom",
        legend.direction = "horizontal",
        legend.justification = "center",
        legend.box.margin = margin(0.5, 0.5, 0.5, 0.5),
        
        plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(size = 12),
        panel.border = element_blank(),
        panel.spacing = unit(1, "lines"))
  
barplot_new
```